### 2021-08-19 MHD/notes.tex
update notes
nonlinear notes and script

\newcommand{\pd}[1]{\partial_{#1}}
\newcommand{\vhat}[1]{\hat{\boldsymbol{#1}}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\M}[1]{\mathbf{#1}}
\newcommand{\grad}{\vec{\nabla}}
\newcommand{\cross}{\vec{\times}}
\newcommand{\curl}{\grad \vec{\times}}
\newcommand{\laplacian}{\nabla^2}
\newcommand{\veclaplacian}{\grad^2}

@@

\section{MRI Notes}
\subsection{Daniel (8/6/2021)}
\subsubsection{Potential Form}
The induction equation is given by
\begin{align*}
    \partial_t\vec{b} &= \grad \cross (\vec{u} \cross \vec{b}) + \eta \grad^2 \vec{b}
    \intertext{Using the following identity}
    \grad \cross \grad \cross \vec{f} &= \grad \grad \cdot \vec{f} - \grad^2\vec{f}
    \intertext{and assumming $\eta$ to be constant, we use $\grad \cdot \vec{b} = 0$, giving}
    \partial_t\vec{b} &= \grad \cross (\vec{u} \cross \vec{b}) - \eta \grad \cross \grad \cross \vec{b}.
    \intertext{Then we define a vector potential $\grad \cross \vec{A} \equiv \vec{b}$, yielding}
    \partial_t \grad \cross \vec{A} &= \grad \cross (\vec{u} \cross \vec{b}) - \grad \cross (\eta \grad \cross \vec{b}) \\
    \partial_t \vec{A} &= \vec{u} \cross \vec{b} - \eta \grad \cross \vec{b} + \grad \phi
    \intertext{where $\phi$ is a scalar potential arizing from ``uncurling'' the equation. We must then provide an additional constraint to fix $\phi$: the Coulomb gauge $\grad \cdot \vec{A} = 0$. Therefore}
    -\grad \cross \vec{b} &= - \curl \grad \cross \vec{A} = \veclaplacian \vec{A}.
\end{align*}
Next we decompose $\vec{u} = \vec{u_0} + \vec{u'}$ and $\vec{b} = \vec{b_0} + \vec{b'}$.
We assume the mean quantities $\vec{u_0}$ and $\vec{b_0}$ are themselves solutions to the original problem. 
If we consider only the 0th mode of $\vec{b}$, i.e. $\vec{b} \cdot \vec{\hat{e}_i} \sim e^{i0}$ then clearly $\veclaplacian \vec{b} = \vec{0}$ and therefore
\begin{align*}
    \partial_t \vec{b'} &= \curl (\vec{u_0} \cross \vec{b'}) + \curl (\vec{u'} \cross \vec{b_0}) + \curl (\vec{u'} \cross \vec{b'}).
    \intertext{Using another identity}
    \curl (\vec{A} \cross \vec{B}) &= \vec{A} \grad \cdot \vec{B} - \vec{B} \grad \cdot \vec{A} + \vec{B} \cdot \grad \vec{A} - \vec{A} \cdot \grad \vec{B}
\end{align*}

\subsection*{Momentum Equation}
Verbatim from Jeff Oishi's MRI paper:
\begin{align*}
    \frac{D\vec{u'}}{Dt} + f\vhat{z} \cross\vec{u'} + Su_x' \vhat{y} + \grad p' + \nu \curl \omega' &= B_0 \partial_z \vec{b'}
    \intertext{where $f$ is the corriolis parameter, S is the background shearing rate, and $B_0 \vhat{z}$ is a uniform background magnetic field. This equation is linearized wrt perturbations. Accordingly, the material derivative goes like}
    \frac{D}{Dt} &\equiv \partial_t + \vec{u} \cdot \grad \\
    &= \partial_t + Sx\partial_y
    \intertext{due to the background velocity $\vec{\overline{u}} = Sx\vhat{y}$. 
    In the nonlinear case we have}
    &= \partial_t + (Sx\vhat{y} + \vec{u'}) \cdot \grad
    \intertext{From inspection and stuff, the irrotational momentum equation goes like}
    \frac{D\vec{u}}{Dt} + \grad p + \nu\cross \omega &= \vec{b} \cdot \grad\vec{b}
    \intertext{Next we generalize $\vec{u} = \vec{u'} + Sx\vhat{y}$ and $\vec{b} = \vec{b'} + B_0\vhat{z}$, giving}
    \underline{\partial_t \vec{u'} + \vec{u'} \cdot \grad \vec{u'} + Sx\partial_y \vec{u'} + Su'_x \vhat{y}} + \grad p + \nu \curl \omega &= B_0 \partial_z\vec{b'} + \vec{b'} \cdot \grad \vec{b'}
    \intertext{where the material derivative $\frac{D\vec{u}}{Dt}$ consists of the underlined terms}
\end{align*}

>>>>

\newcommand{\pd}[1]{\partial_{#1}}
\newcommand{\vhat}[1]{\hat{\boldsymbol{#1}}}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\M}[1]{\mathbf{#1}}
\newcommand{\grad}{\vec{\nabla}}
\newcommand{\cross}{\vec{\times}}
\newcommand{\curl}{\grad \vec{\times}}
\newcommand{\divergence}{\grad \cdot}
\newcommand{\laplacian}{\nabla^2}
\newcommand{\veclaplacian}{\grad^2}

@@

\section*{MRI Notes}
\subsection*{Induction Equation: Potential Form}
The induction equation is given by
\begin{align*}
    \partial_t\vec{b} &= \grad \cross (\vec{u} \cross \vec{b}) + \eta \grad^2 \vec{b}
    \intertext{Using the following identity}
    \grad \cross \grad \cross \vec{f} &= \grad \grad \cdot \vec{f} - \grad^2\vec{f}
    \intertext{and assumming $\eta$ to be constant, we use $\grad \cdot \vec{b} = 0$, giving}
    \partial_t\vec{b} &= \grad \cross (\vec{u} \cross \vec{b}) - \eta \grad \cross \grad \cross \vec{b}.
    \intertext{Then we define a vector potential $\grad \cross \vec{A} \equiv \vec{b}$, yielding}
    \partial_t \grad \cross \vec{A} &= \grad \cross (\vec{u} \cross \vec{b}) - \grad \cross (\eta \grad \cross \vec{b}) \\
    \partial_t \vec{A} &= \vec{u} \cross \vec{b} - \eta \grad \cross \vec{b} + \grad \phi
    \intertext{where $\phi$ is a scalar potential arizing from ``uncurling'' the equation. We must then provide an additional constraint to fix $\phi$: the Coulomb gauge $\grad \cdot \vec{A} = 0$. Therefore}
    -\grad \cross \vec{b} &= - \curl \grad \cross \vec{A} = \veclaplacian \vec{A}.
\end{align*}
Next we decompose $\vec{u} = \vec{u_0} + \vec{u'}$ and $\vec{b} = \vec{b_0} + \vec{b'}$.
We assume the mean quantities $\vec{u_0}$ and $\vec{b_0}$ are themselves solutions to the original problem. 
If we consider only the 0th mode of $\vec{b}$, i.e. $\vec{b} \cdot \vec{\hat{e}_i} \sim e^{i0}$ then clearly $\veclaplacian \vec{b} = \vec{0}$ and therefore
\begin{align*}
    \partial_t \vec{b'} &= \curl (\vec{u_0} \cross \vec{b'}) + \curl (\vec{u'} \cross \vec{b_0}) + \curl (\vec{u'} \cross \vec{b'}).
    \intertext{Using another identity}
    \curl (\vec{A} \cross \vec{B}) &= \vec{A} \grad \cdot \vec{B} - \vec{B} \grad \cdot \vec{A} + \vec{B} \cdot \grad \vec{A} - \vec{A} \cdot \grad \vec{B}
\end{align*}

\subsection*{Momentum Equation: Nonlinear Terms}
Verbatim from Jeff Oishi's MRI paper:
\begin{align*}
    \frac{D\vec{u'}}{Dt} + f\vhat{z} \cross\vec{u'} + Su_x' \vhat{y} + \grad p' + \nu \curl \vec{\omega}' &= B_0 \partial_z \vec{b'}
    \intertext{where $f$ is the corriolis parameter, S is the background shearing rate, and $B_0 \vhat{z}$ is a uniform background magnetic field. This equation is linearized wrt perturbations. Accordingly, the material derivative goes like}
    \frac{D}{Dt} &\equiv \partial_t + \vec{u} \cdot \grad \\
    &= \partial_t + Sx\partial_y
    \intertext{due to the background velocity $\vec{\overline{u}} = Sx\vhat{y}$. 
    In the nonlinear case we have}
    &= \partial_t + (Sx\vhat{y} + \vec{u'}) \cdot \grad
    \intertext{From inspection and stuff, the irrotational momentum equation goes like}
    \frac{D\vec{u}}{Dt} + \grad p + \nu\cross \vec{\omega} &= \vec{b} \cdot \grad\vec{b}
    \intertext{Next we generalize $\vec{u} = \vec{u'} + Sx\vhat{y}$ and $\vec{b} = \vec{b'} + B_0\vhat{z}$, giving}
    \underline{\partial_t \vec{u'} + \vec{u'} \cdot \grad \vec{u'} + Sx\partial_y \vec{u'} + Su'_x \vhat{y}} + \grad p + \nu \curl \vec{\omega} &= B_0 \partial_z\vec{b'} + \vec{b'} \cdot \grad \vec{b'}
    \intertext{where the material derivative $\frac{D\vec{u}}{Dt}$ consists of the underlined terms}
\end{align*}

\subsection*{Induction Equation: Nonlinear Terms}
The MHD induction equation (as given by \textit{Fluid Mechanics of Planets and Stars, 2019}) is given by
\begin{align*}
    \partial_t \vec{b} &= \underline{\curl (\vec{u} \cross \vec{b})} + \eta \laplacian \vec{b}.
    \intertext{We expand the underlined term using the following identity}
    \curl(\vec{u}\cross\vec{b}) &= \vec{u}\divergence\vec{b} - \vec{b}\divergence\vec{u} + \vec{b}\cdot\grad\vec{u} - \vec{u}\cdot\grad\vec{b}
    \intertext{where the first two terms on the RHS vanish due to incompressibility and the abscence of magnetic monopoles. 
    Therefore}
    \partial_t \vec{b} + \vec{u} \cdot \grad\vec{b} &= \vec{b}\cdot\grad\vec{u} + \eta\laplacian\vec{b}.
    \intertext{Substituting the decompositions for $\vec{u}$ and $\vec{b}$ as above yields}
    \partial_t (B_0\vhat{z} + \vec{b'}) + (Sx\vhat{y} + \vec{u'}) \cdot \grad(B_0\vhat{z} + \vec{b'}) &= (B_0\vhat{z} + \vec{b'})\cdot\grad(Sx\vhat{y} + \vec{u'}) + \eta\laplacian(B_0\vhat{z} + \vec{b'}) \\
    \partial_t \vec{b'} + (Sx\vhat{y} + \vec{u'}) \cdot \grad \vec{b'} &= (B_0\vhat{z} + \vec{b'})\cdot\grad(Sx\vhat{y} + \vec{u'}) + \eta\laplacian\vec{b'} 
    \intertext{dropping the $'$s and taking the $x,y,$ and $z$ components of the above yields}
    \partial_t b_x + Sx \partial_y b_x + \vec{u} \cdot \grad b_x &= B_0\partial_z u_x + \vec{b} \cdot \grad u_x + \eta\laplacian b_x \\
    \partial_t b_y + Sx \partial_y b_y + \vec{u} \cdot \grad b_y &= B_0\partial_z u_y + \vec{b} \cdot \grad u_y + S b_x + \eta\laplacian b_y \\
    \partial_t b_z + Sx \partial_y b_z + \vec{u} \cdot \grad b_z &= B_0\partial_z u_z + \vec{b} \cdot \grad u_z + \eta\laplacian b_z 
    \intertext{Note that the linear operator on the LHS is identical for each scalar equation. Accordingly, we redefine $D_t$ to be this operator, i.e.}
    D_tA &\equiv \partial_t A + Sx \partial_y A + \vec{u} \cdot \grad A.
    \intertext{Recall that the current density is given by}
    \vec{j} = j_x\vhat{x} + j_y\vhat{y} + j_z\vhat{z} &= \curl \vec{b} \\
    j_x &= \partial_y b_z - \partial_z b_y
    \intertext{We then rewrite the laplacian term as follows}
    \eta\laplacian \vec{b} &= -\curl\curl \vec{b} = -\curl \vec{j}.
    \intertext{Therefore the $x$-component of the induction equation is given by}
    D_t b_x - B_0\partial_z u_x + \eta(\partial_yj_z - \partial_zj_y) &= \vec{b} \cdot \grad u_x \\
    \intertext{Finally, we combine the $x$ and $y$ components of the induction equation. Taking the material derivative of the $x$-component of the current density gives}
    D_t j_x &= \partial_y D_t b_z - \partial_z D_t b_y.
\end{align*}
\begin{align*}
    \intertext{We have equations for that stuff!}
    D_t j_x &= \partial_y (B_0\partial_z u_z + \vec{b} \cdot \grad u_z + \eta\laplacian b_z) - \partial_z (B_0\partial_z u_y + \vec{b} \cdot \grad u_y + S b_x + \eta\laplacian b_y) \\
    &= B_0 \partial_z \omega_x + \vec{b} \cdot \grad \omega_x + \eta\laplacian j_x - S\partial_z b_x
\end{align*}
Rearranging and expanding gives\\

$\partial_t j_x + Sx\partial_y j_x - B_0 \partial_z \omega_x + S\partial_z b_x - \eta\laplacian j_x = \vec{b} \cdot \grad \omega_x - \vec{u} \cdot \grad j_x$
